Automated discovery of interpretable gravitational-wave population models

### 2022-07-11 src/tex/ms.tex
Last part remain

Internal to \pysr, a multi-objective optimization problem is solved: the evolution seeks to find expressions which are as accurate as possible, while being as simple as possible.
Formally, for our setting, \pysr solves the following optimization problem:
\begin{align}
    s_c &= 
        \mathop{\arg\min}_{
            \tiny
            \begin{array}{c}s\in\mathcal{S}\\ \abs{s} = c \end{array}
        }
        \mathcal{L}(s),
        \\
    &\text{for each } c \in \{1, \ldots, 80\},\nonumber\\
    \text{where }\mathcal{L}(s) &= \sum_i \ell(s(\mathbf{x}_i), y_i)
\end{align}
where $s$ is an expression, 
$\mathcal{S}$ is a space of possible expressions considered,
$\abs{s}$ is the complexity of an expression $s$,
$\ell$ is the loss function, and $\{(\mathbf x_i, y_i)\}$ is the dataset
for $\mathbf x_i$ each input and $y_i$ the label.
The most accurate expression at each $c$, denoted as $s_c$, is computed for a given set range of $c$, which we choose as $c\in\{1, \ldots, 80\}$.
Complexity is defined here
as a sum of the complexity of each symbol in the expression,
and $\ell(\hat{y}_i, y_i) = \qty(\hat{y}_i - y_i)^2$. \todo{What complexities were chosen?}

@@

\todo{Note how much computational power was used, what other PySR settings were used. Could link the script in the paper repo.}

@@

Here, we use symbolic regression for the first time to learn population density models in astrophysics.
One of the main advantages of symbolic regression is to learn a model into an interpretable language, and, therefore, we define a set of operators which frequently occur in GW population models: addition, subtraction, division, power laws, Gaussians (unary operator: $f(x)=\exp(-x^2)$), and conditional statements (binary operator: $g(x, y)=\operatorname{if}(x \geq 0, y, 0)$). \todo{Specify how combinations of operators are ruled out}

@@

GW data comes in the form of time series, in which each time sample carries very little physical meaning.
It is easier to understand the physical implication of a GW event through its inferred source properties \cite{Veitch:2014wba}.
In the case of a BBH, the source properties include the masses of the two merging black holes.
Understanding the population of GW events supports out understanding of the distribution of source properties, such as the mass distribution of black holes \cite{2019PASA...36...10T,Vitale:2020aaz}.

@@

We show the workflow of this work in figure \ref{fig:Gaussian_m1_fit}.
On the leftmost panel, we show the posterior density in primary mass for every event in GWTC3.
In the middle panel \todo{Shouldn’t it be left-middle?}, we show a GMM fitted to the data given by \cite{Tiwari:2020vym}.
Given the \ppd shown in the middle panel, we use symbolic regression to extract effective descriptions of the data. 
As we increase the complexity of the equation, it is natural to expect that the equation should fit the data better than a lower complexity equation since it has additional free parameters.
At the same time, an equation with higher complexity could either be overfitting or difficult to interpret.
\pysr uses a score (see \citealt{cranmerDiscoveringSymbolicModels2020} for details) as a metric normalized to complexity to show how well an equation is fitting the data.
At each complexity, \pysr outputs the equation with the lowest loss.
To select the equations shown in figure \ref{fig:Gaussian_m1_fit}, we first make an accuracy cut based on the loss of the function, so the selected equation would be a good fit to the data.
Then under a certain loss level, we select the equation that has the highest score.
We choose three accuracy levels to select three different equations in our analysis \todo{Specify how the thresholds were chosen}.
The equations selected under these criteria are tabulated in table \ref{tab:equations}.
Note that we take the absolute value of the equations in order to make figure \ref{fig:Gaussian_m1_fit}.\todo{Is this still true?}

@@

% Brief summary of the paper
In this paper, we use symbolic regression to extract interpretable equations from the \ppd of a flexible model fitted to the catalog of GWs.
We show SR can discover equations which are similar to state-of-the-art phenomenological models.
On top of finding forms that have already been discovered,
our result also shed insights on an open question---whether there exist additional features other than the two prominent peaks in the primary mass distribution \cite{Tiwari:2021yvr}.
Our results suggest interpreting any additional features such as tapering of the continuum or wiggles between the two peaks can be tricky as their significance are unclear. \todo{Clean up language}

@@

In this work, we focus on analyzing only the primary mass distribution.
Our method is trivially generalizable to higher-dimensional data, such as the joint space of both the primary mass and the secondary mass.
So far there are relatively few phenomenological models which can capture correlations between observables,
mainly due to the difficulty of manually constructing reasonable models to account for these correlations.
At the same time, the most advanced flexible model in the literature is an n-dimensional histogram, which is hard to interpret.
Our method can be a natural way to explore and to explain these correlations. \todo{Say something about how this can be used for other astrophysical phenomena!}

>>>>

Internal to \pysr, a multi-objective optimization problem is solved: the evolution seeks to find expressions which are as accurate as possible, while being as simple as possible.
Formally, for our setting, \pysr solves the following optimization problem:
\begin{align}
    s_c &= 
        \mathop{\arg\min}_{
            \tiny
            \begin{array}{c}s\in\mathcal{S}\\ \abs{s} = c \end{array}
        }
        \mathcal{L}(s),
        \\
    &\text{for each } c \in \{1, \ldots, 80\},\nonumber\\
    \text{where }\mathcal{L}(s) &= \sum_i \ell(s(\mathbf{x}_i), y_i)
\end{align}
where $s$ is an expression, 
$\mathcal{S}$ is a space of possible expressions considered,
$\abs{s}$ is the complexity of an expression $s$,
$\ell$ is the loss function, and $\{(\mathbf x_i, y_i)\}$ is the dataset
for $\mathbf x_i$ each input and $y_i$ the label.
The most accurate expression at each $c$, denoted as $s_c$, is computed for a given set range of $c$, which we choose as $c\in\{1, \ldots, 80\}$.
Complexity is defined here
as a sum of the complexity of each symbol in the expression,
and $\ell(\hat{y}_i, y_i) = \qty(\hat{y}_i - y_i)^2$.
% \todo{What complexities were chosen?}

@@

% \todo{Note how much computational power was used, what other PySR settings were used. Could link the script in the paper repo.}

@@

Here, we use symbolic regression for the first time to learn population density models in astrophysics.
One of the main advantages of symbolic regression is to learn a model into an interpretable language, and, therefore, we define a set of operators which frequently occur in GW population models: addition, subtraction, division, power laws, Gaussians (unary operator: $f(x)=\exp(-x^2)$), and conditional statements (binary operator: $g(x, y)=\operatorname{if}(x \geq 0, y, 0)$). 
% \todo{Specify how combinations of operators are ruled out}

@@

GW data comes in the form of time series, in which each time sample carries very little physical meaning.
It is easier to understand the physical implication of a GW event through its inferred source properties \cite{Veitch:2014wba}.
In the case of a BBH, the source properties include the masses of the two merging black holes.
On top of extracting physics from individual events,
we can also learn about astrophysical processes through understanding of the distribution of source properties, such as the mass distribution of black holes \cite{2019PASA...36...10T,Vitale:2020aaz}.

@@

We show the workflow of this work in figure \ref{fig:Gaussian_m1_fit}.
On the leftmost panel, we show the posterior density in primary mass for every event in GWTC3.
In the left-middle panel, we show a GMM fitted to the data given by \cite{Tiwari:2020vym}.
Given the \ppd shown in the middle panel, we use symbolic regression to extract effective descriptions of the data. 
As we increase the complexity of the equation, it is natural to expect that the equation should fit the data better than a lower complexity equation since it has additional free parameters.
At the same time, an equation with higher complexity could either be overfitting or difficult to interpret.
\pysr uses a score (see \citealt{cranmerDiscoveringSymbolicModels2020} for details) as a metric normalized to complexity to show how well an equation is fitting the data.
At each complexity, \pysr outputs the equation with the lowest loss.
To select the equations shown in figure \ref{fig:Gaussian_m1_fit}, we first make an accuracy cut based on the loss of the function, so the selected equation would be a good fit to the data.
Then under a certain loss level, we select the equation that has the highest score.
We choose three accuracy levels to select three different equations in our analysis.
% \todo{Specify how the thresholds were chosen}
The equations selected under these criteria are tabulated in table \ref{tab:equations}.
Note that we take the absolute value of the equations in order to make figure \ref{fig:Gaussian_m1_fit}.

@@

% Brief summary of the paper
In this paper, we use symbolic regression to extract interpretable equations from the \ppd of a flexible model fitted to the catalog of GWs.
We show SR can discover equations which are similar to state-of-the-art phenomenological models.
On top of finding forms that have already been discovered,
our result also shed insights on an open question---whether there exist additional features other than the two prominent peaks in the primary mass distribution \cite{Tiwari:2021yvr}.
The additional suggest interpreting any additional features such as tapering of the continuum or wiggles between the two peaks can be tricky as their significance are unclear.

@@

In this work, we focus on analyzing only the primary mass distribution.
Our method is trivially generalizable to higher-dimensional data, such as the joint space of both the primary mass and the secondary mass.
So far there are relatively few phenomenological models which can capture correlations between observables,
mainly due to the difficulty of manually constructing reasonable models to account for these correlations.
At the same time, the most advanced flexible model in the literature is an n-dimensional histogram, which is hard to interpret.
Our method can be a natural way to explore and to explain these correlations.
